Synthesising Recursive Functions for First-Order Model Counting:\\ Challenges, Progress, and Conjectures

### 2023-03-07 doc/paper/paper.tex
Formatting improvements

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand{\Done}{\mathsf{D}_{1}}
\newcommand{\Dtwo}{\mathsf{D}_{2}}
\newcommand{\Dthree}{\mathsf{D}_{3}}
\newcommand{\predicate}{\texttt{\textup{p}}}

\DeclareMathOperator{\friends}{\texttt{friends}}
\DeclareMathOperator{\smokes}{\texttt{smokes}}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\Preds}{Preds}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. The development of
lifted inference algorithms coincided with work on probabilistic relational
models that combine the syntactic power of first-order logic with probabilistic
models such as Bayesian and Markov networks, allowing for a more relational view
of uncertainty
modelling~\cite{DBLP:series/synthesis/2016Raedt,DBLP:journals/ml/KimmigMG15,DBLP:journals/ml/RichardsonD06}.
Lifted inference techniques for probabilistic databases, while developed
somewhat independently, have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC has received more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21,DBLP:journals/corr/abs-2302-04606}.
There is also a recent line of work that uses FOMC-based methods to count
solutions to constraint satisfaction
problems~\cite{DBLP:journals/jair/TotisDRK23}.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{DBLP:conf/starai/JaegerB12}. Over more than a decade, many
classes of instances were shown to be liftable. First, Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class of all
sentences of first-order logic with up to two variables (denoted \FOtwo{}) is
liftable. Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that
there exists a sentence with three variables for which FOMC is
$\#\P_{1}$-complete (i.e., \FOthree{} is not liftable). Since these two seminal
results, most of the research on (W)FOMC focused on developing faster solutions
for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}, and
\Ctwo{} extended with axioms describing trees~\cite{DBLP:conf/kr/BremenK21},
directed acyclic graphs~\cite{DBLP:journals/corr/abs-2302-09830}, and linear
order~\cite{DBLP:journals/corr/abs-2211-01164}. On the empirical front, there
are several implementations of exact WFOMC algorithms:
\textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16},
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21},
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{Alchemy}, \textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20},
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12},
\textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}, and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can
translate sentences in a variant of function-free many-sorted first-order logic
with equality to this internal format. We use lowercase Latin letters for
predicates (e.g., $\texttt{p}$) and constants (e.g., $x$), uppercase Latin
letters for variables (e.g., $X$), and uppercase Greek letters for domains
(e.g., $\Delta$). Sometimes we write predicate $\texttt{p}$ as $\texttt{p}/n$,
where $n \in \mathbb{N}^{+}$ is the \emph{arity} of $\texttt{p}$. An \emph{atom}
is $\texttt{p}(t_1, \dots, t_n)$ for some predicate $\texttt{p}/n$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg \texttt{p}(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all
relevant (finite) domains. Initially, $\mathcal{D}$ contains all domains
mentioned by the input formula. During compilation, new domains are added to
$\mathcal{D}$ by some of the compilation rules. Each such new domain is
interpreted as a subset of some other domain in $\mathcal{D}$.

@@

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $\texttt{p}/2$, variables $X$, $Y$, $Z$, and domains
  $\Gamma$ and $\Delta$. All variables that occur as the first argument to
  $\texttt{p}$ are in $\Gamma$, and, likewise, all variables that occur as the
  second argument to $\texttt{p}$ are in $\Delta$. Therefore, $\phi$ `type
  checks' as a valid formula.
\end{example}

One can read a formula in our format as a sentence in first-order logic. All
variables in a clause are implicitly universally quantified, and all clauses in
a formula are implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{equation}\label[formula]{eq:formula}
  \begin{split}
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
      Y \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(X, Z)) \land{}
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
      X \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(Z, Y)).
    \end{multlined}
  \end{split}
\end{equation}

@@

\begin{example}\label{example:simple}
  Let $\Delta$ be a domain of size $n \in \mathbb{N}_{0}$. The model count of
  $\forall X \in \Delta\text{. } \texttt{p}(X) \lor \texttt{q}(X)$ is then
  $3^{n}$. Intuitively, since both predicates are of arity one, they can be
  interpreted as subsets of $\Delta$. Thus, the formula says that each element
  of $\Delta$ has to be included in $\texttt{p}$ or in $\texttt{q}$ or in both,
  i.e., there are three possibilities.
\end{example}

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\Gamma = \Delta = \{\,1, 2\,\}$, i.e., $\sigma(\Gamma) = \sigma(\Delta) = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $\texttt{p}$. The empty relation does. All four relations of
  cardinality one (e.g., $\{\, (1, 1) \,\}$) do too. Finally, there are two
  relations of cardinality two---$\{\, (1, 1), (2, 2) \,\}$ and
  $\{\, (1, 2), (2, 1) \,\}$---that satisfy the conditions as well. Thus, the
  FOMC of $(\phi, \sigma)$ is 7. Incidentally, the FOMC of $\phi$ counts partial
  injections from $\Gamma$ to $\Delta$. We will continue to use the problem of
  counting partial injections (and the formula from \cref{example:first}
  specifically) as the main running example.
\end{example}

@@

\begin{example}\label{example:dr}
  Let $\phi$ be a formula with a single clause
  \begin{multline*}
    (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}).
  \end{multline*}
  Then we can introduce constant $x \in \Gamma$ and rewrite $\phi$ as
  $\phi' = \{\, c_{1}, c_{2} \,\}$, where
  \begin{align*}
    c_{1} &= \begin{multlined}[t]
      (\{\, \neg \texttt{p}(x, Y), \neg \texttt{p}(x, Z) \,\}, \{\, (Y, Z) \,\}, \\
      \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
      \end{multlined}\\
    c_{2} &= \begin{multlined}[t]
      (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
      \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Delta \,\}),
      \end{multlined}
  \end{align*}
  and $\Gamma' = \Gamma \setminus \{\, x \,\}$.
\end{example}

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in evidently
  unsatisfiable formulas. We impose two restrictions on $W$. First,
  $W^2 \cap E = \emptyset$ ensures that there are no pairs of variables in $W$
  that are constrained to be distinct, since that would result in an $x \ne x$
  constraint after substitution. Similarly, we want to avoid variables in $W$
  that have inequality constraints with constants: after the substitution, such
  constraints would transform into inequality constraints between two constants.
  In this case, both subsets of $V$ satisfy these conditions, and
  \cref{line:generation} generates two clauses for the output formula:
  \begin{multline*}
    (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg \texttt{p}(x, Y), \neg \texttt{p}(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

  When \cref{line:forclause} picks $c = c_2$, then $V = \{\, X, Z \,\}$. The
  subset $W = V$ fails to satisfy the conditions on \cref{line:conditions}
  because of the $X \ne Z$ constraint. The other three subsets of $V$ all
  generate clauses for $\phi'$. Indeed, $W = \emptyset$ generates
  \begin{multline*}
    (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(Z, Y) \,\}, \{\, (X, Z), (X, x), (Z, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\}),
  \end{multline*}
  $W = \{\, X \,\}$ generates
  \[
    (\{\, \neg \texttt{p}(x, Y), \neg \texttt{p}(Z, Y) \,\}, \{\, (Z, x) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Gamma \,\}),
  \]
  and $W = \{\, Z \,\}$ generates
  \[
    (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(x, Y) \,\}, \{\, (X, x) \,\}, \{\, X \mapsto \Gamma, Y \mapsto \Delta \,\}).
  \]
\end{example}

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &=
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2')$, where
  \begin{align*}
    c_1' &=
           \begin{multlined}[t]
             (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Delta \,\}),
           \end{multlined} \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

  \ForEach{formula and node $(\psi, v) \in C(\#\phi)$}{\label{line:selectformula}
    $\rho \gets \identifyRecursion{$\phi$, $\psi$}$\;
    \lIf{$\rho \ne {\normalfont \texttt{null}}$}{\Return{$\{\, (\Reff_\rho(v), \langle\rangle) \,\}$}}\label{line:rho}
  }
  \Return{$\emptyset$}\;

>>>>

\SetKwProg{Fn}{Function}{:}{}

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand{\Done}{\domainNames_{1}}
\newcommand{\Dtwo}{\domainNames_{2}}
\newcommand{\Dthree}{\domainNames_{3}}
\newcommand{\predicate}{\texttt{\textup{p}}}
\newcommand{\predicateq}{\texttt{\textup{q}}}
\newcommand{\nulll}{\texttt{\textup{null}}}

\DeclareMathOperator{\friends}{\texttt{\textup{friends}}}
\DeclareMathOperator{\smokes}{\texttt{\textup{smokes}}}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\Preds}{Preds}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. The development of
lifted inference algorithms coincided with work on probabilistic relational
models that combine the syntactic power of first-order logic with probabilistic
models such as Bayesian and Markov networks, allowing for a more relational view
of uncertainty
modelling~\cite{DBLP:series/synthesis/2016Raedt,DBLP:journals/ml/KimmigMG15,DBLP:journals/ml/RichardsonD06}.
Lifted inference techniques for probabilistic databases, while developed
somewhat independently, have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC has received more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.
There is also a recent line of work that uses FOMC-based methods to count
solutions to constraint satisfaction
problems~\cite{DBLP:journals/jair/TotisDRK23}.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{DBLP:conf/starai/JaegerB12}. Over more than a decade, many
classes of instances were shown to be liftable. First, Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class of all
sentences of first-order logic with up to two variables (denoted \FOtwo{}) is
liftable. Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that
there exists a sentence with three variables for which FOMC is
$\#\P_{1}$-complete (i.e., \FOthree{} is not liftable). Since these two seminal
results, most of the research on (W)FOMC focused on developing faster solutions
for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}, and
\Ctwo{} extended with axioms for trees~\cite{DBLP:conf/kr/BremenK21}. On the
empirical front, there are several implementations of exact WFOMC algorithms:
\textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16},
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21},
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{Alchemy}, \textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20},
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12},
\textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}, and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can
translate sentences in a variant of function-free many-sorted first-order logic
with equality to this internal format. We use lowercase Latin letters for
predicates (e.g., $\predicate$) and constants (e.g., $x$), uppercase Latin
letters for variables (e.g., $X$), and uppercase Greek letters for domains
(e.g., $\Delta$). Sometimes we write predicate $\predicate$ as $\predicate/n$,
where $n \in \mathbb{N}^{+}$ is the \emph{arity} of $\predicate$. An \emph{atom}
is $\predicate(t_1, \dots, t_n)$ for some predicate $\predicate/n$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg \predicate(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all
relevant (finite) domains. Initially, $\mathcal{D}$ contains all domains
mentioned by the input formula. During compilation, new domains are added to
$\mathcal{D}$ by some of the compilation rules. Each such new domain is
interpreted as a subset of some other domain in $\mathcal{D}$.

@@

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \predicate(X, Y), \neg \predicate(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $\predicate/2$, variables $X$, $Y$, $Z$, and domains
  $\Gamma$ and $\Delta$. All variables that occur as the first argument to
  $\predicate$ are in $\Gamma$, and, likewise, all variables that occur as the
  second argument to $\predicate$ are in $\Delta$. Therefore, $\phi$ `type
  checks' as a valid formula.
\end{example}

One can read a formula in our format as a sentence in first-order logic. All
variables in a clause are implicitly universally quantified, and all clauses in
a formula are implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{equation}\label[formula]{eq:formula}
  \begin{split}
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
      Y \ne Z \Rightarrow \neg \predicate(X, Y) \lor \neg \predicate(X, Z)) \land{}
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
      X \ne Z \Rightarrow \neg \predicate(X, Y) \lor \neg \predicate(Z, Y)).
    \end{multlined}
  \end{split}
\end{equation}

@@

\begin{example}\label{example:simple}
  Let $\Delta$ be a domain of size $n \in \mathbb{N}_{0}$. The model count of
  $\forall X \in \Delta\text{. } \predicate(X) \lor \predicateq(X)$ is then
  $3^{n}$. Intuitively, since both predicates are of arity one, they can be
  interpreted as subsets of $\Delta$. Thus, the formula says that each element
  of $\Delta$ has to be included in $\predicate$ or in $\predicateq$ or in both,
  i.e., there are three possibilities.
\end{example}

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\Gamma = \Delta = \{\,1, 2\,\}$, i.e., $\sigma(\Gamma) = \sigma(\Delta) = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $\predicate$. The empty relation does. All four relations of
  cardinality one (e.g., $\{\, (1, 1) \,\}$) do too. Finally, there are two
  relations of cardinality two---$\{\, (1, 1), (2, 2) \,\}$ and
  $\{\, (1, 2), (2, 1) \,\}$---that satisfy the conditions as well. Thus, the
  FOMC of $(\phi, \sigma)$ is 7. Incidentally, the FOMC of $\phi$ counts partial
  injections from $\Gamma$ to $\Delta$. We will continue to use the problem of
  counting partial injections (and the formula from \cref{example:first}
  specifically) as the main running example.
\end{example}

@@

\begin{example}\label{example:dr}
  Let $\phi$ be a formula with a single clause
  \begin{multline*}
    (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}).
  \end{multline*}
  Then we can introduce constant $x \in \Gamma$ and rewrite $\phi$ as
  $\phi' = \{\, c_{1}, c_{2} \,\}$, where
  \begin{align*}
    c_{1} &= \begin{multlined}[t]
      (\{\, \neg \predicate(x, Y), \neg \predicate(x, Z) \,\}, \{\, (Y, Z) \,\}, \\
      \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
      \end{multlined}\\
    c_{2} &= \begin{multlined}[t]
      (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
      \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Delta \,\}),
      \end{multlined}
  \end{align*}
  and $\Gamma' = \Gamma \setminus \{\, x \,\}$.
\end{example}

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in evidently
  unsatisfiable formulas. We impose two restrictions on $W$. First,
  $W^2 \cap E = \emptyset$ ensures that there are no pairs of variables in $W$
  that are constrained to be distinct, since that would result in an $x \ne x$
  constraint after substitution. Similarly, we want to avoid variables in $W$
  that have inequality constraints with constants: after the substitution, such
  constraints would transform into inequality constraints between two constants.
  In this case, both subsets of $V$ satisfy these conditions, and
  \cref{line:generation} generates two clauses for the output formula:
  \begin{multline*}
    (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg \predicate(x, Y), \neg \predicate(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

  When \cref{line:forclause} picks $c = c_2$, then $V = \{\, X, Z \,\}$. The
  subset $W = V$ fails to satisfy the conditions on \cref{line:conditions}
  because of the $X \ne Z$ constraint. The other three subsets of $V$ all
  generate clauses for $\phi'$. Indeed, $W = \emptyset$ generates
  \begin{multline*}
    (\{\, \neg \predicate(X, Y), \neg \predicate(Z, Y) \,\}, \{\, (X, Z), (X, x), (Z, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\}),
  \end{multline*}
  $W = \{\, X \,\}$ generates
  \[
    (\{\, \neg \predicate(x, Y), \neg \predicate(Z, Y) \,\}, \{\, (Z, x) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Gamma \,\}),
  \]
  and $W = \{\, Z \,\}$ generates
  \[
    (\{\, \neg \predicate(X, Y), \neg \predicate(x, Y) \,\}, \{\, (X, x) \,\}, \{\, X \mapsto \Gamma, Y \mapsto \Delta \,\}).
  \]
\end{example}

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &=
          \begin{multlined}[t]
            (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg \predicate(X, Y), \neg \predicate(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2')$, where
  \begin{align*}
    c_1' &=
           \begin{multlined}[t]
             (\{\, \neg \predicate(X, Y), \neg \predicate(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Delta \,\}),
           \end{multlined} \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg \predicate(X, Y), \neg \predicate(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

  \ForEach{formula and node $(\psi, v) \in C(\#\phi)$}{\label{line:selectformula}
    $\rho \gets \identifyRecursion{$\phi$, $\psi$}$\;
    \lIf{$\rho \ne \nulll$}{\Return{$\{\, (\Reff_\rho(v), \langle\rangle) \,\}$}}\label{line:rho}
  }
  \Return{$\emptyset$}\;

  \Fn{\identifyRecursion{formula $\phi$, formula $\psi$, map $\rho = \emptyset$}}{
    \lIf{$|\phi| \ne |\psi|$ {\bf or} $\#\phi \ne \#\psi$}{\Return{\nulll}}
    \lIf{$\phi = \emptyset$}{\Return{$\rho$}}
    \ForEach{clause $c \in \psi$\label{line:for1}}{
      \ForEach{clause $d \in \phi$ s.t. $\#d=\#c$\label{line:for2}}{
        \ForAll{$\gamma \in \generateMaps{$c$, $d$, $\rho$}$\label{line:generateMaps}}{
          $\rho' \gets \identifyRecursion{$\phi\setminus\{\, d \,\}$, $\psi\setminus\{\, c \,\}$, $\rho\cup\gamma$}$\;\label{line:recursion}
          \lIf{$\rho' \ne \nulll$}{\Return{$\rho'$}}
        }
      }
      \Return{\nulll}\;
    }
  }
\end{algorithm}
